---
title: "Write title here"

author: 
  - Sevnur Kulak
  - Dongqi Bi
  - Ivan Li
thanks: "Code and data are available at: https://github.com/hellowoshibb/STA304_paper3"
date: "`r format(Sys.time(), '%d %B %Y')`"
abstract: "write abstract here"
output: 
  bookdown::pdf_document2:
    fig_caption: yes
bibliography: references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(hash)
library(dplyr)
library(forcats)
library(knitr)
library(janitor)
library(tidyverse)
library(tidyr)
library(wrapr)
library(haven)
library(ggplot2)
library(gridExtra)
```

```{r, message=FALSE, cache=TRUE, warning = FALSE, echo = FALSE}
#read gss raw data
gss <- read_csv(here::here("input/data/gss.csv"))
```

```{r, age-first-child, fig.cap="Age of respondents having first child ", warning=FALSE, echo=FALSE, message = FALSE}

gss_afb <- gss |>
  select(age_at_first_birth, sex) |>
  na.omit()


gss_afb |>
  ggplot(mapping = aes(x = age_at_first_birth))+
  geom_histogram(fill="steelblue3", colour = "white", aes(y = (..count..)/sum(..count..)))+
  labs(title = "Distribution of Age of Respondents having their First Child",
       y = "Proportion of respondents",
       fill = "Proportion of respondents")

```



```{r, female-age-first-child, fig.cap="Distribution of Age of respondents having their first child by gender ", message = FALSE, warning=FALSE, echo=FALSE}

gss_afb <- gss |>
  select(age_at_first_birth, sex) |>
  na.omit()

p1 <- gss_afb |>
  filter(sex == "Female") |>
  ggplot(mapping = aes(x = age_at_first_birth))+
  geom_histogram(aes(y = (..count..)/sum(..count..)), fill="steelblue3", colour = "white")+
  labs(title = "Distribution of Age of Females",
       y = "Proportion of respondents",
       fill = "Proportion of respondents")+
  theme(plot.title = element_text(size=10))


  
p2 <- gss_afb |>
  filter(sex == "Male") |>
  ggplot(mapping = aes(x = age_at_first_birth))+
  geom_histogram(aes(y = (..count..)/sum(..count..)), fill="steelblue3", colour = "white")+
  labs(title = "Distribution of Age of Males",
       y = "Proportion of respondents",
       fill = "Proportion of respondents")+
  theme(plot.title = element_text(size=10))

cowplot::plot_grid(p1, p2)

```

```{r, education, fig.cap="Distribution of Age at first birth by education levels ", warning=FALSE, echo=FALSE, message = FALSE}

ed <- gss |>
  select(age_at_first_birth, education)|>
  filter(education == "High school diploma or a high school equivalency certificate" |
         education == "Less than high school diploma or its equivalent" |
         education == "Bachelor's degree (e.g. B.A., B.Sc., LL.B.)"|
        education == "University certificate, diploma or degree above the bach...")
  
ed[ed == "High school diploma or a high school equivalency certificate"] <- "High School"
ed[ed == "Less than high school diploma or its equivalent"] <- "Less than high achool"
ed[ed == "Bachelor's degree (e.g. B.A., B.Sc., LL.B.)"] <- "Bachelor"
ed[ed == "University certificate, diploma or degree above the bach..."] <- "Above Bachelor"

ed |>
  ggplot(mapping = aes(x = age_at_first_birth, fill=education))+
  geom_histogram(aes(y = (..count..)/sum(..count..)))+
  labs(title = "Distribution of Age at first birth by education levels" ,,
       y = "Counts of respondents",
       fill = "Education levels")
```



```{r, occupation, fig.cap="Distribution of Age at first birth by occupation ", warning=FALSE, echo=FALSE, message = FALSE}


oc <- gss |>
  select(age_at_first_birth, occupation, income_respondent)|>
  filter(occupation == "Business, finance, and administration occupations" |
         occupation == "Sales and service occupations" |
         occupation == "Health occupations"|
         occupation == "Management occupations")
  
oc[oc == "Business, finance, and administration occupations"] <- "Business occupation"
oc[oc == "Sales and service occupations"] <- "Sales occupation"
oc[oc == "Management occupations"] <- "Management occupations"

oc %>%
  mutate(income_respondent = fct_relevel(income_respondent, 
            "Less than $25,000", "$25,000 to $49,999", "$50,000 to $74,999", "$75,000 to $99,999","$100,000 to $ 124,999", "$125,000 and more" ))%>%
  ggplot(aes(x=income_respondent, y=age_at_first_birth, color = factor(occupation))) +
  geom_point()+
  labs(title = "Distribution of Age at first birth by occupation" ,,
       color = "Occupation")
#oc %>%
  #ggplot(aes(income_respondent, age_at_first_birth, color = factor(income_respondent)))+
  #geom_point()




```

```{r, income_comparison, fig.cap="Distribution of Age at first birth by income", warning=FALSE, echo=FALSE, message = FALSE}

p3 <- oc |>
  filter(income_respondent == "Less than $25,000") |>
  ggplot(mapping = aes(x = age_at_first_birth))+
  geom_histogram(aes(y = (..count..)/sum(..count..)), fill="steelblue3", colour = "white")+
  labs(title = "income<$25K",
       y = "Proportion of respondents",
       fill = "Proportion of respondents")+
  theme(plot.title = element_text(size=10))


p4 <- oc |>
  filter(income_respondent == "$125,000 and more") |>
  ggplot(mapping = aes(x = age_at_first_birth))+
  geom_histogram(aes(y = (..count..)/sum(..count..)), fill="steelblue3", colour = "white")+
  labs(title = "income<$125K",
       y = "Proportion of respondents",
       fill = "Proportion of respondents")+
  theme(plot.title = element_text(size=10))

cowplot::plot_grid(p3, p4)
```

<<<<<<< HEAD
```{r, feelings_in_life, fig.cap="Number of total children", warning=FALSE, echo=FALSE, message = FALSE}

gss_fl <- gss %>%
  select(age_at_first_birth, feelings_life) %>%
  na.omit()

gss_fl <- gss_fl %>%
  group_by(feelings_life) %>%
  summarize(mean(age_at_first_birth))%>%
  rename(mean_age = `mean(age_at_first_birth)`)
  

gss_fl %>% 
  ggplot(aes(feelings_life, mean_age)) +
  geom_point() +
  geom_line()+
  labs(title = "Age of having first child vs feelings about life as a whole")

```
=======
```{r, incomeboxplot, fig.cap="Distribution of Age at first birth (income>$125K)", warning=FALSE, echo=FALSE}

oc <- gss |>
  select(age_at_first_birth, occupation, income_respondent)|>
  filter(occupation == "Business, finance, and administration occupations" |
         occupation == "Sales and service occupations" |
         occupation == "Health occupations"|
         occupation == "Management occupations")
  
oc[oc == "Business, finance, and administration occupations"] <- "Business occupation"
oc[oc == "Sales and service occupations"] <- "Sales occupation"
oc[oc == "Management occupations"] <- "Management occupations"


p1 = oc %>% mutate(income_respondent = fct_relevel(income_respondent, 
            "Less than $25,000", "$25,000 to $49,999", "$50,000 to $74,999", "$75,000 to $99,999","$100,000 to $ 124,999", "$125,000 and more" )) |> filter(occupation == "Health occupations") |>  ggplot(aes(x=income_respondent, y=age_at_first_birth)) +
  geom_boxplot(colour = 'green')+  geom_jitter(alpha = 0.3, width = 0.05, height = 0, colour = 'green')+
  labs(title = "Health Occupation Distribution" ,,
       color = "Occupation")+
  theme(axis.text.x = element_text(angle = 35)) + labs(x='Income', y='Age at First Birth')

p2 = oc %>% mutate(income_respondent = fct_relevel(income_respondent, 
            "Less than $25,000", "$25,000 to $49,999", "$50,000 to $74,999", "$75,000 to $99,999","$100,000 to $ 124,999", "$125,000 and more" )) |> filter(occupation == "Business occupation") |>  ggplot(aes(x=income_respondent, y=age_at_first_birth)) +
  geom_boxplot(colour = 'yellow')+  geom_jitter(alpha = 0.3, width = 0.05, height = 0, colour = 'yellow')+
  labs(title = "Business Occupation Distribution" ,,
       color = "Occupation")+
  theme(axis.text.x = element_text(angle = 35)) + labs(x='Income', y='Age at First Birth')

p3 = oc %>% mutate(income_respondent = fct_relevel(income_respondent, 
            "Less than $25,000", "$25,000 to $49,999", "$50,000 to $74,999", "$75,000 to $99,999","$100,000 to $ 124,999", "$125,000 and more" )) |> filter(occupation == "Sales occupation") |>  ggplot(aes(x=income_respondent, y=age_at_first_birth)) +
  geom_boxplot(colour = 'blue')+  geom_jitter(alpha = 0.3, width = 0.05, height = 0, colour = 'blue')+
  labs(title = "Sales Occupation Distribution" ,,
       color = "Occupation")+
  theme(axis.text.x = element_text(angle = 35)) + labs(x='Income', y='Age at First Birth')

p4 = oc %>% mutate(income_respondent = fct_relevel(income_respondent, 
            "Less than $25,000", "$25,000 to $49,999", "$50,000 to $74,999", "$75,000 to $99,999","$100,000 to $ 124,999", "$125,000 and more" )) |> filter(occupation == "Management occupations") |>  ggplot(aes(x=income_respondent, y=age_at_first_birth)) +
  geom_boxplot(colour = 'red')+  geom_jitter(alpha=0.3, width = 0.05, height = 0, colour = 'red')+
  labs(title = "Management Occupation Distribution" ,,
       color = "Occupation")+
  theme(axis.text.x = element_text(angle = 35)) + labs(x='Income', y='Age at First Birth')

cowplot::plot_grid(p1, p2, p3, p4)
```
>>>>>>> 0ea02d63d58847abd8699a582d805da51d807d9d
